Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-21 co-regular-EPQ.tex
first co-regular

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ be a data-differentiated execution. Then, $e$ is not $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is not $\Gamma$-linearizable, for {\color{blue}any} maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch) 
{\color {blue}We prove an equivalent property: $e$ is $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.} %We consider the case of $\mathsf{MatchedMaxPriority}$ and the case of $\mathsf{UnmatchedMaxPriority}$ is similar.

To prove the $\textit{only if}$ direction, let $e$ be a data-differentiated execution linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$ imposes no restriction on the operations in $u$, $v$ and $w$ with priorities incomparable to $p$, erasing all these operations results in a sequential execution which still satisfies this property. Similarly, for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

@@

A differentiated execution $e$ is not $\mathsf{MatchedMaxPriority}$-linearizable when all the $\textit{put}$ operations in $e$ using the maximal priority $p$ are matched, and $e$ is not linearizable w.r.t. the set of sequential executions satisfying $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ for {\color {blue}any value $x$ of priority $p$}. We consider two cases depending on whether $e$ contains exactly one value with priority $p$ or at least two values. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with the condition that all the values other than $x$ have a priority strictly smaller than $p$ (corresponding to the first case), and by $\mathsf{MatchedMaxPriority}^{=}$ the strengthening of the same formula with the negation of this condition (corresponding to the second case).
We use particular instances of register automata~\cite{finite-memory, the paper I found} whose states include only two registers, one for storing a priority guessed at the initial state, and one for storing the priority of the current action in the execution. The transitions can check equality or the order relation $\prec$ between the values stored in the two registers. Instead of formalizing the full class of register automata, we consider a simpler class which suffices our needs. More precisely, we consider a class of labeled transition systems whose states consist of a finite control part and a register $r$ interpreted to elements of $\mathbb{P}$. The transition labels can be one of the following:
\begin{itemize}
	\item $r=*$ for storing an arbitrary value to $r$,
	\item $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},a)$ for reading call/return actions of a remove,
	\item $\textit{call}(\textit{put},d,g)$ where $g\in\{=r,\prec r,true\}$ is a guard, for reading a call action $\textit{call}(\textit{put},d,p)$ of a put and checking whether $p$ is either equal to or smaller than the value stored in $r$, or arbitrary,
	\item $\textit{ret}(\textit{put},d,true)$ for reading a return action $\textit{ret}(\textit{put},d,p)$ for some $p$.
\end{itemize}
The set of words accepted by such a transition system can be defined as usual.

@@

%To facilitate the proof of co-Regular of $\mathsf{MatchedMaxPriority}$, we separate $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ into the following two sub-predicates, where the first one represents that $x$ is the only value with priority $p$ in $e$ , and the second one represents that there are more than one value with priority $p$ in $e$. Here $\textit{occurOnce}(p,l)$ holds if there is only one value with priority $p$ in $l$.
%
%{\small
%\begin{align*}
%\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(e,x)=\mathsf{true}$,} \\
%&\hspace{4mm}\mbox{$\textit{occurOnce}(p,e)$ where $p$ is the priority of $x$} \\
%\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(e,x)=\mathsf{true}$,} \\
%&\hspace{4mm}\mbox{$\neg\textit{occurOnce}(p,e)$ where $p$ is the priority of $x$} \\
%\end{align*}
%}
%
%$\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ is obtained from $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Conc}(e,x)$ by using $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)$ instead of $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$. And similarly for $\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Conc}(e,x)$.
%
%We can also define two conditions, where $\textit{unmatched-priorities}(l)$ is the set of priorities occurring in $\textit{put}$ operations of $e$ for which there is $\textit{rm}$ operation removing the same value:
%
%{\small
%\begin{align*}
%\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{matched-priorities}(e)$ for a maximal priority} \\
%&\hspace{4mm}\mbox{$p\in priorities(e)$, $\textit{occurOnce}(p,e)$} \\
%\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Condition}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{matched-priorities}(e)$ for a maximal priority} \\
%&\hspace{4mm}\mbox{$p\in priorities(e)$, $\neg\textit{occurOnce}(p,e)$}
%\end{align*}}
We give a typical example of an execution $e$ which is not $\mathsf{MatchedMaxPriority}^>$-linearizable in
%Given a data-differentiated execution $e$ such that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)$ holds, Lemma \ref{lemma:automata for extended priority queue with single priority} is not enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$. This is because that it is possible that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ does not holds because of interaction between actions of multiple priorities. One such example is shown in
\figurename~\ref{fig:introduce gap for EPQ1Lar}. Intuitively, this is a violation because during the whole execution of $\textit{rm}(b)$, the priority queue stores a smaller priority value (which should be removed before $b$). To be more precise, we define \emph{the interval of a value $x$} as the time interval from the return of a put $\textit{ret}(\textit{put},x,p)$ to the call of the matching remove $\textit{call}(rm,x)$, or to the end of the execution if such a call action doesn't exist. Intuitively, it represents the time interval in which a value is guaranteed to be stored into the concurrent priority queue.
In \figurename~\ref{fig:introduce gap for EPQ1Lar}, we draw the interval of each value by dashed line. Here we assume that $p_1 \prec p_4$, $p_2 \prec p_4$ and $p_3 \prec p_4$. We can not find a sequence $s$ where $e \sqsubseteq s$ and $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,b)$ holds, since each time point from $\textit{call}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is included in the interval of some smaller priority value, and $\textit{rm}(b)$ can not happen in interval of a smaller priority value.

@@

TODO GIVE THE FORMAL DEFINITION OF THE LEFT-RIGHT CONSTRAINT
{\color {blue}Let us introduce left-right constraint to detect the phenomenon in \figurename~\ref{fig:introduce gap for EPQ1Lar}.

\begin{definition}\label{def:left-right constraint for matched put and rm operations}
Given a data-differentiated $e$ without $\textit{rm}(\textit{empty})$, and let $\textit{put}(x,p)$ and $\textit{rm}(x)$ be the only operations with priority $p$. The left-right constraint of $\textit{put}(x,p)$ and $\textit{rm}(x)$ is the graph $G$ where:

\begin{itemize}
\item the nodes are the values of $e$, to which we add a node,
\item there is an edge from $d_1$ to $x$, if $\textit{put}(d_1,\_) <_{\textit{hb}} \textit{put}(x,p)$ or $\textit{rm}(x)$,
\item there is an edge from $x$ to $d_1$, if $\textit{rm}(x)$ happens before $\textit{rm}(d_1)$ or $\textit{rm}(d_1)$ does not exists,
\item there is an edge from $d_1$ to $d_2$, if $\textit{put}(d_1,\_) <_{\textit{hb}} \textit{rm}(d_2,\_)$.
\end{itemize} 
\end{definition}

The existence of cycle though value with maximal priority in left-right constraint formally modelled the phenomenon in \figurename~\ref{fig:introduce gap for EPQ1Lar}. For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$ is a cycle in left-right constraint. We then prove that getting rid of cycle though value with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$.
}

When there is a cycle $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ through value $x$ with maximal priority in left-right constraint, we say that $x$ is covered by $d_1,\ldots,d_m$. Assume that $x$ is covered by $d_1,\ldots,d_m$, we can safely rename $x$ into $b$, rename $d_1,\ldots,d_m$ into $a$ and rename all other value into $c$ by data-independence. Such execution can be recognized by register automata, since between the first $\textit{ret}(\textit{put},a,\_)$ and the last $\textit{call}(\textit{rm},a)$ (if exists), $\textit{ret}(\textit{put},a,\_)$ and $\textit{call}(\textit{rm},a)$ occurs in pair and there is no need for count. In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we construct a set $\textit{Auts}_{\textit{1-lar}}$ of register automata, {\color {blue}and prove that $\textit{Auts}_{\textit{1-lar}}$ is $\mathsf{MatchedMaxPriority}^{>}$-complete, as stated by the following lemma.

>>>>

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ a data-differentiated execution. Then, $e$ is $\Gamma$-linearizable iff $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch) 
%{\color {blue}We prove an equivalent property: $e$ is $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.} %We consider the case of $\mathsf{MatchedMaxPriority}$ and the case of $\mathsf{UnmatchedMaxPriority}$ is similar.
To prove the $\textit{only if}$ direction, let $e$ be a data-differentiated execution linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$ imposes no restriction on the operations in $u$, $v$ and $w$ with priorities incomparable to $p$, erasing all these operations results in a sequential execution which still satisfies this property. Similarly, for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

@@

A differentiated execution $e$ is not $\mathsf{MatchedMaxPriority}$-linearizable when all the $\textit{put}$ operations in $e$ using the maximal priority $p$ are matched, and $e$ is not linearizable w.r.t. the set of sequential executions satisfying $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ for each value $x$ of priority $p$. We consider two cases depending on whether $e$ contains exactly one value with priority $p$ or at least two values. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with the condition that all the values other than $x$ have a priority strictly smaller than $p$ (corresponding to the first case), and by $\mathsf{MatchedMaxPriority}^{=}$ the strengthening of the same formula with the negation of this condition (corresponding to the second case).
We use particular instances of register automata~\cite{finite-memory, the paper I found} whose states include only two registers, one for storing a priority guessed at the initial state, and one for storing the priority of the current action in the execution. The transitions can check equality or the order relation $\prec$ between the values stored in the two registers. Instead of formalizing the full class of register automata, we consider a simpler class which suffices our needs. More precisely, we consider a class of labeled transition systems whose states consist of a finite control part and a register $r$ interpreted to elements of $\mathbb{P}$. The transition labels can be one of the following:
\begin{itemize}
	\item $r=*$ for storing an arbitrary value to $r$,
	\item $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},a)$ for reading call/return actions of a remove,
	\item $\textit{call}(\textit{put},d,g)$ where $g\in\{=r,\prec r,true\}$ is a guard, for reading a call action $\textit{call}(\textit{put},d,p)$ of a put and checking whether $p$ is either equal to or smaller than the value stored in $r$, or arbitrary,
	\item $\textit{ret}(\textit{put},d,true)$ for reading a return action $\textit{ret}(\textit{put},d,p)$ for some $p$.
\end{itemize}
The set of words accepted by such a transition system can be defined as usual.

@@

%To facilitate the proof of co-Regular of $\mathsf{MatchedMaxPriority}$, we separate $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ into the following two sub-predicates, where the first one represents that $x$ is the only value with priority $p$ in $e$ , and the second one represents that there are more than one value with priority $p$ in $e$. Here $\textit{occurOnce}(p,l)$ holds if there is only one value with priority $p$ in $l$.
%
%{\small
%\begin{align*}
%\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(e,x)=\mathsf{true}$,} \\
%&\hspace{4mm}\mbox{$\textit{occurOnce}(p,e)$ where $p$ is the priority of $x$} \\
%\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(e,x)=\mathsf{true}$,} \\
%&\hspace{4mm}\mbox{$\neg\textit{occurOnce}(p,e)$ where $p$ is the priority of $x$} \\
%\end{align*}
%}
%
%$\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ is obtained from $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Conc}(e,x)$ by using $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)$ instead of $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$. And similarly for $\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Conc}(e,x)$.
%
%We can also define two conditions, where $\textit{unmatched-priorities}(l)$ is the set of priorities occurring in $\textit{put}$ operations of $e$ for which there is $\textit{rm}$ operation removing the same value:
%
%{\small
%\begin{align*}
%\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{matched-priorities}(e)$ for a maximal priority} \\
%&\hspace{4mm}\mbox{$p\in priorities(e)$, $\textit{occurOnce}(p,e)$} \\
%\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Condition}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{matched-priorities}(e)$ for a maximal priority} \\
%&\hspace{4mm}\mbox{$p\in priorities(e)$, $\neg\textit{occurOnce}(p,e)$}
%\end{align*}}
We give a typical example of an execution $e$ which is not $\mathsf{MatchedMaxPriority}^>$-linearizable in
%Given a data-differentiated execution $e$ such that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)$ holds, Lemma \ref{lemma:automata for extended priority queue with single priority} is not enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$. This is because that it is possible that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ does not holds because of interaction between actions of multiple priorities. One such example is shown in
\figurename~\ref{fig:introduce gap for EPQ1Lar}. Intuitively, this is a violation because during the whole execution of $\textit{rm}(b)$, the priority queue stores a smaller priority value (which should be removed before $b$). To be more precise, we define \emph{the interval of a value $x$} as the time interval from the return of a put $\textit{ret}(\textit{put},x,p)$ to the call of the matching remove $\textit{call}(rm,x)$, or to the end of the execution if such a call action doesn't exist. Intuitively, it represents the time interval in which a value is guaranteed to be stored into the concurrent priority queue.
In \figurename~\ref{fig:introduce gap for EPQ1Lar}, we draw the interval of each value by dashed line. Here we assume that $p_1 \prec p_4$, $p_2 \prec p_4$, and $p_3 \prec p_4$. We can not find a sequence $s$ where $e \sqsubseteq s$ and $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,b)$ holds, since each time point from $\textit{call}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is included in the interval of some smaller priority value, and $\textit{rm}(b)$ can't take effect in the interval of a smaller priority value.

@@

To formalize the scenario in \figurename~\ref{fig:introduce gap for EPQ1Lar} we use the notion of \emph{left-right constraint} defined below.

\begin{definition}\label{def:left-right constraint for matched put and rm operations}
Let $e$ be a data-differentiated execution which contains only one maximal priority $p$, and only one value $x$ of priority $p$ (and no $\textit{rm}(\textit{empty})$ operations). 
% and let $\textit{put}(x,p)$ and $\textit{rm}(x)$ be the only operations with priority $p$. 
The \emph{left-right constraint of $x$} is the graph $G$ where: %$\textit{put}(x,p)$ and $\textit{rm}(x)$ is the graph $G$ where:
\begin{itemize}
\item the nodes are the values occurring in $e$, %, to which we add a node,
\item there is an edge from $d_1$ to $x$, if $\textit{put}(d_1,\_) <_{\textit{hb}} \textit{put}(x,p)$ or $\textit{put}(d_1,\_) <_{\textit{hb}} \textit{rm}(x)$,
\item there is an edge from $x$ to $d_1$, if $\textit{rm}(x)<_{\textit{hb}}\textit{rm}(d_1)$ or $\textit{rm}(d_1)$ does not exists,
\item there is an edge from $d_1$ to $d_2$, if $\textit{put}(d_1,\_) <_{\textit{hb}} \textit{rm}(d_2,\_)$.
\end{itemize} 
\end{definition}

The execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} is not $\mathsf{MatchedMaxPriority}^>$-linearizable because the left-right constraint of the maximal priority value $b$ contains a cycle: $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$. We prove that the presence of such a cycle is equivalent to the execution not being $\mathsf{MatchedMaxPriority}^>$-linearizable (see Appendix~\ref{}).
%The existence of cycle though value with maximal priority in left-right constraint formally modelled the phenomenon in \figurename~\ref{fig:introduce gap for EPQ1Lar}. For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$ is a cycle in left-right constraint. We then prove that getting rid of cycle though value with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$.
When the left-right constraint of the maximal priority value $x$ contains a cycle of the form $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ for some $d_1$,$\ldots$,$d_n\in \mathbb{D}$, we say that $x$ is \emph{covered} by $d_1,\ldots,d_m$. The shape of such a cycle can be detected using our class of automata, the only complication being the unbounded number of values $d_1$,$\ldots$,$d_n$. However, by data independence, whenever an implementation contains such an execution it also contains an execution where all the values $d_1$,$\ldots$,$d_n$ are renamed to the same value $a$, and $x$ is renamed to $b$. Therefore, our automata can be defined over a fixed set of values $a$, $b$, and $\top$ (for the operations outside of the non-linearizable projection).

TODO WHY IT IS A SET OF AUTOMATA ? SAY WHAT ARE THE CASES THAT NEED TO BE COVERED. SAY WHAT IS THE CASE WE GIVE IN THE FIGURE. USE $\top$ AS MENTIONED AT THE BEGINNING OF THE SECTION INSTEAD OF 	$c$. USE BIG CAPS FOR SETS OF TRANSITION LABELS.

Assume that $x$ is covered by $d_1,\ldots,d_m$, we can safely rename $x$ into $b$, rename $d_1,\ldots,d_m$ into $a$ and rename all other value into $c$ by data-independence. Such execution can be recognized by register automata, since between the first $\textit{ret}(\textit{put},a,\_)$ and the last $\textit{call}(\textit{rm},a)$ (if exists), $\textit{ret}(\textit{put},a,\_)$ and $\textit{call}(\textit{rm},a)$ occurs in pair and there is no need for count. In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we construct a set $\textit{Auts}_{\textit{1-lar}}$ of register automata, {\color {blue}and prove that $\textit{Auts}_{\textit{1-lar}}$ is $\mathsf{MatchedMaxPriority}^{>}$-complete, as stated by the following lemma.
